Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-31 article-models-tt.tex
added stuff to the intro of section 4

\section{Operational syntactics for type theory before type constructors}\label{tt}
In this section we give a description of dependent type theory before type
constructors with explicit operations of extension, weakening and substitution. We will
formulate the type theory in such a way that contexts aren't defined
to be lists of variable
declarations. We rather take a syntactic approach in our treatment of contexts,
axiomatizing only what you can do with them and allowing for the freedom to
interpret the axiom in any desired way. 
The main reason is that we don't see variable declarations in the internal
models either. This way we also set out to a more algebraic approach of type
theory and higher category theory. Thirdly, we will not have to be burdened with
superficial comments about variables being bounded or not, or fresh or free or 
not occuring at all.

@@

To get the overview of the compatibility rules, we list the sections
where these compatibility rules are described in the following table (in this
table, the
subsection mentioned in row $X$ and column $Y$ consideres the rules of the
operation $Y\circ X$):
\begin{center}
\begin{tabular}{r|ccc}
& extension & weakening & substitution\\
\hline
extension & \autoref{comp-ee} & \autoref{comp-ew} & \\
weakening & \autoref{comp-we} & \autoref{comp-ww} & \autoref{comp-ws}\\
substitution & \autoref{comp-se} & \autoref{comp-sw} & \autoref{comp-ss}
\end{tabular}
\end{center}
Since extension by a context acts trivially on families and terms there are
no rules added for extension followed by substitution.

In \autoref{extension-on-terms} we come back to the operation of extension, but
now for terms. The rules in this section explain what it means to be a term of
an extended family, namely to be an extended term. The compatibility rules for
extended terms are then similar as the compatibility rules that have appeared
earlier: there is a compatibility rule for the extension of an extension, for
the weakening of an extended term, for the substutution of an extended term by
an arbitrary term and for substitution by an extended term. The last set of
compatibility rules is a form of currying-uncurrying -- this famous
correspondence naturally comes out of our scheme of compatibility rules.

\subsection{Judgments and inference rules}
The type theory we describe here is a theory of contexts, families of
contexts and terms thereof. The families of contexts are by some authors called
dependent contexts, but they are handled a bit differently here because they
become the primary object of study. Dependent contexts can be types; they could
be seen as atomic or indecomposable dependent contexts.

@@

\subsection{Composition and identity terms}
\subsubsection{The defining property of weakening}
The judgmental equalities we're about to describe assert that substituting a term
in the weakening a thing gives you the thing back. In the case of contexts we get that each fiber
$\subst{x}{\ctxwk{A}{B}}$ is just $B$ and in the case of terms we get 
that $\ctxwk{A}{y}$ is the constant function
mapping everything to $y:B$. Thus, these rules actually establish the weakening
as the weakening. After stating the rules we will describe what it means to
compose context morphisms (terms of weakened contexts).

@@

\subsection{Digging a little deeper in the theory of contexts, families and terms}

>>>>

\section{The theory of contexts, families and terms}\label{tt}
In this section we give a description of dependent type theory before type
constructors with explicit operations of extension, weakening and substitution. We will
formulate the type theory in such a way that contexts aren't defined
to be lists of variable
declarations. We rather take a syntactic approach in our treatment of contexts,
axiomatizing only what you can do with them and allowing for the freedom to
interpret the axiom in any desired way. 
The main reason is that we don't see variable declarations in the internal
models either. This way we also set out to a more algebraic approach of type
theory and higher category theory. Thirdly, we will not have to be burdened with
superficial comments about variables being bounded or not, or fresh or free or 
not occuring at all.

@@

To get the overview of the compatibility rules, we list the sections
where these compatibility rules are described in the following table (in this
table, the
subsection mentioned in row $X$ and column $Y$ consideres the rules of the
operation $Y\circ X$):
\begin{center}
\begin{tabular}{r|ccc}
& extension & weakening & substitution\\
\hline
extension & \autoref{comp-ee} & \autoref{comp-ew} & \\
weakening & \autoref{comp-we} & \autoref{comp-ww} & \autoref{comp-ws}\\
substitution & \autoref{comp-se} & \autoref{comp-sw} & \autoref{comp-ss}
\end{tabular}
\end{center}

In \autoref{digging_deeper} we use the framework we have developed in this
section to derive new notions and their properties. In particular, we will
develop the notion of \emph{extension on terms} together with the projection
maps from an extension to the `base' context of family, the 
\emph{family pullback} which is a version of pullbacks for families and thirdly
the \emph{inductive morphisms} which are morphisms of type theory that allow
to find terms of families over the codomain context by pulling them back to
the domain context and finding a term there.

In \autoref{extension-on-terms} on the extension operation on terms we will
derive all the compatibility rules that one would expect to hold for extension
on terms. The key to most of these results is the currying operation, which
could be seen as the missing feature in the table above. The extension on terms
operation depends in an essential way on the substitution operation, on the
identity terms and therefore indirectly also on the weakening operation. Thus,
we will see here all of the features of the theory we develop in
\autoref{judgments} through \autoref{categorical_properties} come to the 
surface.

It should be kept in mind though that in the current formulation there is no
sealed deal establishing a relationship between families over a context
and any kind of morphisms -- neither with morphisms having the `base' of the
family as its codomain nor with families into a universe (universes will be
introduced in \autoref{universes}). The only thing we know here is
that a family $P$ over $\ctxext{\Gamma}{A}$ determines a context morphism
from $\ctxext{A}{P}$ to $A$ in context $\Gamma$, the projection. 
We do not see this as
a shortcoming of the theory of contexts families and terms. Rather, such a
correspondence is a feature of a theory which does incorporate universes. The
fact that we're lacking a clear connection between families and (a specified
class of) morphisms, however, does show up in our treatment of the notion we
called familie pullbacks. For instance, we can't show that a square of families
is a pullback precisely when the corresponding square of projections is a
pullback: only the backwards direction holds. Neither can we show that the
composition of two family pullback squares is again a family pullback square.
We feel that pointing out what we can and can't do in the current setting is
an important aspect of developing an intuition with the system and therefore
we include this subsubsection even though the theory might feel `incomplete'
compared to the usual theory of pullbacks.

In the last subsubsection we give a treatment of inductive morphisms. These
morphisms appear also in the introduction of the many inductive type
constructors in \autoref{tt_constructors} and therefore a general treatment of
the subject is insightful.

\subsection{Judgments and inference rules}\label{judgments}
The type theory we describe here is a theory of contexts, families of
contexts and terms thereof. The families of contexts are by some authors called
dependent contexts, but they are handled a bit differently here because they
become the primary object of study. Dependent contexts can be types; they could
be seen as atomic or indecomposable dependent contexts.

@@

\subsection{Composition and identity terms}\label{categorical_properties}
\subsubsection{The defining property of weakening}
The judgmental equalities we're about to describe assert that substituting a term
in the weakening a thing gives you the thing back. In the case of contexts we get that each fiber
$\subst{x}{\ctxwk{A}{B}}$ is just $B$ and in the case of terms we get 
that $\ctxwk{A}{y}$ is the constant function
mapping everything to $y:B$. Thus, these rules actually establish the weakening
as the weakening. After stating the rules we will describe what it means to
compose context morphisms (terms of weakened contexts).

@@

\subsection{Digging a little deeper in the theory of contexts, families and terms}
\label{digging_deeper}

### 2014-03-31 article-models-ttconstr.tex
added stuff to the intro of section 4

\section{Type constructors in structural type theory}
In this section we will introduce the usual type constructors to structural
type theory. We do so with the point of view that each type constructor is a
(class of) operations on type theory that should be compatible with extension,
weakening, substitution and identity morphisms. Likewise, each induction
principle of an inductively defined type constructor is going to be such an
operator, and it is therefore also required to be compatible with extension,
weakening, substitution and identity functions.

@@

\subsection{Universes}
The type operator $\UU$ has as input a pair $(A,P)$ consisting of a family
$A$ in context $\Gamma$ and a family $P$ in context $\ctxext{\Gamma}{A}$. Its
output is a pair $(U(A,P),\tilde{U}(A,P))$ consisting of a family
$U(A,P)$ in context $\Gamma$ and a family $\tilde{U}(A,P)$ over it for which
there are terms representing the empty context, the operations of extension,
weakening and substitution and the identity terms. Moreover there is a morphism
$i$ from $A$ to $U(A,P)$ with the property that 
$\jcomp{}{i}{\tilde{U}(A,P)}\jdeq P$.

>>>>

\section{Type constructors in structural type theory}
\label{tt_constructors}
In this section we will introduce the usual type constructors to structural
type theory. We do so with the point of view that each type constructor is a
(class of) operations on type theory that should be compatible with extension,
weakening, substitution and identity morphisms. Likewise, each induction
principle of an inductively defined type constructor is going to be such an
operator, and it is therefore also required to be compatible with extension,
weakening, substitution and identity functions.

@@

\subsection{Universes}\label{universes}
The type operator $\UU$ has as input a pair $(A,P)$ consisting of a family
$A$ in context $\Gamma$ and a family $P$ in context $\ctxext{\Gamma}{A}$. Its
output is a pair $(U(A,P),\tilde{U}(A,P))$ consisting of a family
$U(A,P)$ in context $\Gamma$ and a family $\tilde{U}(A,P)$ over it for which
there are terms representing the empty context, the operations of extension,
weakening and substitution and the identity terms. Moreover there is a morphism
$i$ from $A$ to $U(A,P)$ with the property that 
$\jcomp{}{i}{\tilde{U}(A,P)}\jdeq P$.
